Title

### 2020-06-21 IWST/paper.tex
shortening abstract, intro and related work

\begin{document}
\title{Title}
\author{Myroslava Romaniuk}
\affiliation{\institution{Ukrainian Catholic University} \city{Lviv} \country{Ukraine}}
\email{romaniuk@ucu.edu.ua}
\begin{abstract}
Code completion is one of the essential features of any IDE and it significantly improves the developer experience and productivity. Good code completion can both speed up the development process, as well as aid the developer in API exploration. On the other hand, having a slow or inaccurate completion can be very cumbersome. Thus, it is important to find a way to make it as effective as possible.

The current implementation of code completion in the Pharo IDE is based on the abstract syntax tree (AST) of source code. The AST allows us to learn the semantic role and the kind of tokens (e.g. class name, method name, literal, et cetera) and to suggest contextually relevant completions. However, the current implementation has no efficient way to sort the completion candidates, which means that sometimes the user must scroll through a long list of proposed completions to find the one that they need.

In this work, I study how code completion can be improved by sorting completions in such a way that most frequently used tokens appear first. To that end, I implement a sorting plugin based on the n-gram language model.

I use quantitative and qualitative evaluation techniques to compare the performance of the n-gram (unigram and bigram) and alphabetic sorters. As a result of the evaluation, the unigram sorter was shown to have the best performance.

The sorting strategies implemented are part of a Pharo IDE plugin. It is open source and is available at \url{https://github.com/myroslavarm/CompletionSorting}.
\end{abstract}

@@

\section{Introduction}
Code completion is one of the most used features in any IDE. Whether it is being used for improving speed and accuracy of typing or as an API guide, helping developers find their way around libraries, code completion is one of the first things a developer notices as soon as they start coding. The speed with which the results are suggested, as well as their accuracy, is essential, and it is something that can certainly "make or break" the workflow of the developer. Therefore, researchers and software engineers are always trying to find new approaches that can improve code completion quality and make it as effective, as accurate, and as fast\footnote{By "fast" we mean that the completion is quick to give suggestions, by "accurate" that the completion proposes contextually correct results, and by "effective" we mean that the combination of both the aforementioned qualities makes the tool successful in achieving the result, which is to make finding the most suitable completion as easy as possible.} as possible.

The main distinction among the many approaches to code completion would be whether the approach uses semantic context or not. Namely, semantic models are models that use code structure analysis to tailor completions to type and/or semantic role, or, in other words, to give contextually correct and relevant results. Machine learning models can also employ semantic analysis but often treat source code as plain text, with the context not taken into consideration.

Throughout this project, I focused on the code completion in Pharo\footnote{\url{http://pharo.org/}}. Pharo is a dynamically typed programming language inspired by Smalltalk, and the Pharo IDE is an interactive IDE intended for developing in Pharo\footnote{n the rest of this thesis, I use the word "Pharo" to refer to the programming language, and "Pharo IDE" to refer to the development environment.}.

@@

However, currently there is no way to effectively sort the completion candidates. The order of suggestions is significant because with a wrong sorting strategy, desired completions can appear at the bottom of the list. This is unproductive as it requires the developer to scroll down or type more symbols.

In this project, the main goal is to come up with an effective sorting strategy that would show relevant results first.

\subsection{Proposed Approach in a Nutshell}
In this work, I study how code completion can be improved using statistical language models for sorting the completion candidates, on top of an existing semantic completion. In particular, I propose to use n-gram language models, as they have been documented to be used for such a task (\cite{Hind12a}), and I believe this strategy has the potential to improve the relevance of sorting in the Pharo IDE.

To find the most relevant completion suggestions based on their probability of occurring in source code, I implement two sorting strategies based on the unigram and bigram models. These models are trained on a large corpus of source code collected from 50 projects written in Pharo (\cite{Zait20a}).

@@

\subsection{Contributions}
Throughout this project I made the following contributions:
\begin{itemize}
    \item Enhanced the NgramModel\footnote{\url{https://github.com/pharo-ai/NgramModel}} library, which is an open-source library that implements n-gram language models in Pharo (4 accepted pull requests).
    \item Trained several n-gram language models on the tokenised source code of the Pharo programming language.
    \item Used those trained models to build a sorting tool that enhances code completion in the Pharo IDE by sorting the proposed suggestions according to their relevance in the given context.
    \item Made one of the two implemented sorters both effective and fast enough to actually be suitable for developer use. Both implementations can be used in the Pharo IDE after being loaded from the CompletionSorting GitHub repository (\url{https://github.com/myroslavarm/CompletionSorting}).
    \item Proposed a combination of two approaches for evaluating the results of code completion: (1) a quantitative evaluation technique inspired by \cite{Robb08a} and (2) a qualitative evaluation where I perform a case study of several completion scenarios and analyse the results.
    \item Part of this work has been accepted for the Student Research Competition at <Programming> 2020. The conference was supposed to take place in Porto, Portugal in March and was cancelled due to COVID-19. Despite that, the abstract is still up for publication in the ACM Digital Library.
    \item To my knowledge, this is the first implementation of a code completion engine in Smalltalk family of IDEs that uses machine learning for sorting completion candidates.
\end{itemize}

\section{Related Works}
\subsection{Earlier Code Completion Systems}
Standard code completions in IDEs used to only rely on language-specific pattern matching, i.e. sorting completions alphabetically based on the symbols already typed in. \cite{Robb08a}, however, showed that code completion could be improved by using program history (program modifications over time). They managed to get good results by prioritising suggestions from recently modified method bodies, and even better results by using per-session vocabulary (changes of the last hour) and merging it with type-based completion.

According to \cite{Bruc09a}, up until 10 years ago, code completion systems were mainly based on type information, with no contextual analysis. The authors countered that by implementing intelligent code completions that learned from examples and had a significantly better performance in terms of the relevance of suggestions than other then-common implementations. Their solutions included the frequency-based code completion (frequency of use of code), an association rule-based completion, and the Best Matching Neighbours code completion (method calls of the closest source snippet found, using a modified k-nearest neighbours algorithm), which was the main contribution of their paper. The BMN based implementation was integrated into Eclipse and demonstrated promising results. It was later extended by \cite{Prok15a} (see \ref{sec:RelatedWorks-DeepLearning}).

\subsection{Software Naturalness}
In the paper "On The Naturalness of Software", \cite{Hind12a} compared source code to natural languages. They claim that code is even more repetitive, predictable and full of patterns than human languages. In the paper, they also argue that code can be modelled by statistical language models, which can be used to support software engineers. Their approach was based on capturing high-level statistical regularity at the n-gram level by taking \textit{n-1} previous tokens that are already entered into the text buffer, and attempting to guess the next token. Using this model, it is possible to estimate the most probable sequences of tokens and suggest the most relevant code completions to developers.

This work served as a catalyst for the following research using natural language processing (NLP) for source code. For instance, \cite{Tu14a} also learnt that code "has a high degree of localness, where identifiers (e.g. variable names) are repeated often within close distance" (\cite{Alla18a}). Thus, they applied a cache mechanism that assigns higher probability to tokens that have been observed most recently, and improved the results even further.

\cite{Nguy13a} enhanced the state-of-the-art n-gram approach by incorporating semantic information into code tokens, rather than treating them as text -- i.e. annotating each token with its data type and semantic role if available, which allowed them to increase predictability even further.

\subsection{Deep Learning for Code Completion}
In the more recent years, researchers started applying deep learning models such as deep recurrent neural networks (RNN).

For instance, \cite{Hell19a} recorded the results of a case study of 15,000 completions (completion events) for VisualStudio. One of the conclusions they reached is that even though RNNs often outperform n-gram models in typical natural language settings, n-gram models are sometimes a better choice for modelling source code. For example, they state that the deep learner is better at core method invocations but loses on third-party library calls, whereas the n-gram model naturally outperforms it on internal API calls but loses out on the other categories.

\cite{Prok15a} worked on an extensible inference engine for intelligent code completion systems, called PatternBased Bayesian Network (PBN). Eclipse Code Recommenders project adapted the PBN approach for their intelligent call completion. They also tested (evaluating quality, speed and model size) Best Matching Neighbour algorithm, using additional context information for more precise recommendations, and applying clustering techniques to improve model sizes. They conclude that showing the developer hundreds of recommendations may be as ineffective as showing none, and intelligent code completions better target the needs of developers that are unfamiliar with an API.

\cite{Hell17a} introduced a dynamically updatable n-gram model which outperformed both the traditional n-gram models and the deep learning RNN and long short-term memory (LSTM) models.

\cite{Li17a} developed an attention mechanism which exploits the parent-children information of the AST of source code. As correctly predicting out-of-vocabulary (OOV) values in code completion is mainly unsuccessful, the authors implemented a pointer mixture network which either generates a new value through an RNN component or copies an OOV value from local context through a pointer component.

\cite{Rayc14a} implemented a tool called SLANG, which first extracts abstract histories from the data. Then, these histories are fed to a language model such as an n-gram model or recurrent neural network model, which treats the histories as sentences in a natural language and learns probabilities for each sentence, without taking into consideration the AST of the code.

\subsection{Existing Code Completion Systems in IDEs}
The code completion system (Pythia) described by \cite{Svya19a} is part of the Intellicode extension used in the Visual Studio Code IDE to complete Python code. Pythia uses LSTM networks trained on long-range code contexts extracted from abstract syntax trees, which allows it to capture semantics carried by distant nodes and helps rank the method and API recommendations for developers more successfully. Python is a dynamically typed language, so to use type information in Pythia they infer types at runtime based on static analysis of user patterns and add this information into the training sequence.

\cite{Asad14a} developed a tool called CSCC (context-sensitive code completion), which is an example-based completion tool, which leverages contextual information to better support method call completion. CSCC uses tokenisation instead of deep parsing to collect method call usage patterns and requires type information of the receiver object. The CSCC tool is available as an Eclipse plugin for the Java Editor. Execution time-wise the tool showed the results roughly equal to the state-of-the-art. However, in terms of method call recommendation accuracy, it outperformed those approaches, including BMN (the Best Matching Neighbours implementation from \cite{Bruc09a}).

CACHECA is a cache language model-based code completion tool for Eclipse's Java editor which was presented by \cite{Fran15a}; the tool is based on the cache language model described by \cite{Tu14a} that was mentioned earlier in this chapter. According to \cite{Fran15a}, CACHECA greatly enhances Eclipse's built-in engine by incorporating both the corpus and locality statistics, especially when no type information is available.

\subsection{Summary}
Here is a brief overview of the progress made in the area of code completion research and development in recent years:

\begin{itemize}
    \item Code completion tools used to only rely on language-specific pattern matching (i.e. by prefix), but around 10 years ago, the idea of "intelligent" code completion that can learn from examples was popularised.
    \item The software engineering approaches following this idea relied on the frequency of use of code, contextual association, and hierarchical proximity.
    \item From another perspective, source code was likened to natural languages because of its repetitiveness and patterns; the idea that source code can be modelled by statistical language models attracted more machine learning approaches for the task of improving code completion.
    \item Following that, many code completion implementations using machine learning ignored such information as type and semantic meaning, and treated it as regular text, trying to infer relevant predictions from source code history alone.
    \item These days, code completion systems that are actually being used by developers (such as plugins for IDEs) mostly combine both the software engineering approach (taking advantage of semantic and contextual information) and the machine learning one (taking advantage of source code patterns).
\end{itemize}

Another thing worth noting is that the majority of the research experiments mentioned in this chapter have been tested on or applied for only statistically typed languages, predominantly Java and C\# (\cite{Hind12a} also trained on C).

>>>>

\begin{document}
\title{Title}
\author{Myroslava Romaniuk}
\affiliation{\institution{Ukrainian Catholic University} \city{Lviv} \country{Ukraine}}
\email{romaniuk@ucu.edu.ua}
\begin{abstract}
Code completion is one of the essential features of any IDE and it significantly improves the developer experience and productivity. Thus, it is important to find a way to make it as effective as possible.

In this work, I propose to improve code completion by sorting completions in such a way that most frequently used tokens appear first. To that end, I implement a sorting plugin based on the n-gram language model.

I use quantitative and qualitative evaluation techniques to compare the performance of the n-gram (unigram and bigram) and alphabetic sorters. As a result of the evaluation, the unigram sorter was shown to have the best performance.
\end{abstract}

@@

\section{Introduction}
\subsection{Context}
Whether it is being used for improving speed and accuracy of typing or as an API guide, helping developers find their way around libraries, code completion is one of the first things a developer notices as soon as they start coding. The speed with which the results are suggested, as well as their accuracy, is essential, and it is something that can certainly "make or break" the workflow of the developer. Therefore, researchers and software engineers are always trying to find new approaches that can improve code completion quality and make it as effective, as accurate, and as fast\footnote{By "fast" we mean that the completion is quick to give suggestions, by "accurate" that the completion proposes contextually correct results, and by "effective" we mean that the combination of both the aforementioned qualities makes the tool successful in achieving the result, which is to make finding the most suitable completion as easy as possible.} as possible.

Throughout this project, I focused on the code completion in Pharo\footnote{\url{http://pharo.org/}}. Pharo is a dynamically typed programming language inspired by Smalltalk, and the Pharo IDE is an interactive IDE intended for developing in Pharo.

@@

Currently there is no way to effectively sort the completion candidates. The order of suggestions is significant because with a wrong sorting strategy, desired completions can appear at the bottom of the list. This is unproductive as it makes developers scroll down or type more symbols. Thus, in this work, the main goal is to come up with an effective sorting strategy that would show relevant results first.

\subsection{Proposed Approach in a Nutshell}
In this work, I study how code completion can be improved using statistical language models for sorting the completion candidates, on top of an existing semantic completion. In particular, I propose to use n-gram language models, as they have been documented to be used for such a task \cite{Hind12a}, and I believe this strategy has the potential to improve the relevance of sorting in the Pharo IDE.

To find the most relevant completion suggestions based on their probability of occurring in source code, I implement two sorting strategies based on the unigram and bigram models. These models are trained on a large corpus of source code collected from 50 projects written in Pharo \cite{Zait20a}.

@@

\section{Related Works}
\subsection{Earlier Code Completion Systems}
Standard code completions in IDEs used to only rely on language-specific pattern matching, i.e. sorting completions alphabetically based on the symbols already typed in. Robbes et al. \cite{Robb08a}, however, showed that code completion could be improved by using program history (program modifications over time). They managed to get good results by prioritising suggestions from recently modified method bodies, and even better results by using per-session vocabulary (changes of the last hour) and merging it with type-based completion.

Up until 10 years ago, code completion systems were mainly based on type information, with no contextual analysis. Bruch et al. \cite{Bruc09a} countered that by implementing intelligent code completions that learned from examples and had a significantly better performance in terms of the relevance of suggestions than other then-common implementations. Their solutions included the frequency-based code completion (frequency of use of code), an association rule-based completion, and the Best Matching Neighbours code completion (method calls of the closest source snippet found, using a modified k-nearest neighbours algorithm). The BMN based implementation was integrated into Eclipse and demonstrated promising results.

\subsection{Software Naturalness}
Hindle et al. \cite{Hind12a} compared source code to natural languages. They claim that code is even more repetitive, predictable and full of patterns than human languages. In the paper, they also argue that code can be modelled by statistical language models, which can be used to support software engineers. Their approach was based on capturing high-level statistical regularity at the n-gram level by taking \textit{n-1} previous tokens that are already entered into the text buffer, and attempting to guess the next token. Using this model, it is possible to estimate the most probable sequences of tokens and suggest the most relevant code completions to developers. This work served as a catalyst for the following research using natural language processing (NLP) for source code.

For instance, Tu et al. \cite{Tu14a} also learnt that code "has a high degree of localness, where identifiers (e.g. variable names) are repeated often within close distance" \cite{Alla18a}. Thus, they applied a cache mechanism that assigns higher probability to tokens that have been observed most recently, and improved the results even further. Nguyen et al. \cite{Nguy13a} enhanced the state-of-the-art n-gram approach by incorporating semantic information into code tokens, rather than treating them as text -- i.e. annotating each token with its data type and semantic role if available, which allowed them to increase predictability even further.

\subsection{Deep Learning for Code Completion}
In the more recent years, researchers started applying deep learning models such as deep recurrent neural networks (RNN). For instance, Hellendoorn et al. \cite{Hell19a} recorded the results of a case study of 15,000 completions (completion events) for VisualStudio. In comparison to n-gram models, they stated that the deep learner is better at core method invocations but loses on third-party library calls, whereas the n-gram model naturally outperforms it on internal API calls but loses out on the other categories. Proksch et al. \cite{Prok15a} worked on an extensible inference engine for intelligent code completion systems, called PatternBased Bayesian Network (PBN). Eclipse Code Recommenders project adapted the PBN approach for their intelligent call completion.

Hellendoorn et al. \cite{Hell17a} introduced a dynamically updatable n-gram model which outperformed both the traditional n-gram models and the deep learning RNN and long short-term memory (LSTM) models. Li et al. \cite{Li17a} developed an attention mechanism which exploits the parent-children information of the AST of source code. As correctly predicting out-of-vocabulary (OOV) values in code completion is mainly unsuccessful, the authors implemented a pointer mixture network which either generates a new value through an RNN component or copies an OOV value from local context through a pointer component. Raychev et al. \cite{Rayc14a} implemented a tool called SLANG, which first extracts abstract histories from the data. Then, these histories are fed to a language model such as an n-gram model or recurrent neural network model, which treats the histories as sentences in a natural language and learns probabilities for each sentence, without taking into consideration the AST of the code.

\subsection{Existing Code Completion Systems in IDEs}
The code completion system (Pythia) \cite{Svya19a} is part of the Intellicode extension used in the Visual Studio Code IDE to complete Python code. Pythia uses LSTM networks trained on long-range code contexts extracted from abstract syntax trees, which allows it to capture semantics carried by distant nodes and helps rank the method and API recommendations for developers more successfully. Python is a dynamically typed language, so to use type information in Pythia they infer types at runtime based on static analysis of user patterns and add this information into the training sequence.

Asaduzzaman et al. \cite{Asad14a} developed a tool called CSCC (context-sensitive code completion), which is an example-based completion tool, which leverages contextual information to better support method call completion. CSCC uses tokenisation instead of deep parsing to collect method call usage patterns and requires type information of the receiver object. In terms of method call recommendation accuracy, it outperformed state-of-the-art approaches, including BMN \cite{Bruc09a}. The CSCC tool is available as an Eclipse plugin for the Java Editor.

CACHECA is a cache language model-based code completion tool for Eclipse's Java editor \cite{Fran15a}; the tool is based on the cache language model \cite{Tu14a}. According to the authors, CACHECA greatly enhances Eclipse's built-in engine by incorporating both the corpus and locality statistics, especially when no type information is available.

\subsection{Conclusion}
It is worth noting that the majority of the research experiments mentioned in this chapter have been tested on or applied for only statistically typed languages, predominantly Java and C\# (\cite{Hind12a} also trained on C).
